---
title: "Biostat 212a Homework 4"
subtitle: "Due Mar. 4, 2025 @ 11:59PM"
author: "Nathaniel Boyle and 006525153"
date: today
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
engine: knitr
knitr:
  opts_chunk: 
    fig.align: 'center'
    # fig.width: 6
    # fig.height: 4
    message: FALSE
    cache: false
---

## ISL Exercise 8.4.3 (10pts)
Consider the Gini index, classifcation error, and entropy in a simple
classifcation setting with two classes. Create a single plot that displays each 
of these quantities as a function of pˆm1. The x-axis should 
display pˆm1, ranging from 0 to 1, and the y-axis should display the
value of the Gini index, classifcation error, and entropy.

```{r}
# Create p^m1 from 0-1, 0.01 step size
p = seq(0, 1, 0.01)

# Define gini index
gini = p * (1 - p) * 2

# Define entropy
entropy = -(p * log(p) + (1 - p) * log(1 - p))

# Define classification error
class.err = 1 - pmax(p, 1 - p)

# Plot each as a function of p^m1
matplot(p, cbind(gini, entropy, class.err), col = c("red", "green", "blue"),
        xlab = expression(paste(hat(p)[m1])), 
        ylab = "Gini Index, Classification Error, and Entropy")
```
Hint: In a setting with two classes, pˆm1 = 1 − pˆm2. You could make
this plot by hand, but it will be much easier to make in R.


## ISL Exercise 8.4.4 (10pts)

This question relates to the plots in Figure 8.14.
(a) Sketch the tree corresponding to the partition of the predictor
space illustrated in the left-hand panel of Figure 8.14.
The numbers inside the boxes indicate the mean of Y within each region.

(b) Create a diagram similar to the left-hand panel of Figure 8.14,
using the tree illustrated in the right-hand panel of the same
fgure. You should divide up the predictor space into the correct
regions, and indicate the mean for each region.


```{r}
par(xpd = NA)
plot(NA, NA, type = "n", xlim = c(-2, 2), ylim = c(-3, 3), xlab = "X1", ylab = "X2")
# X2 < 1
lines(x = c(-2, 2), y = c(1, 1))
# X1 < 1 with X2 < 1
lines(x = c(1, 1), y = c(-3, 1))
text(x = (-2 + 1)/2, y = -1, labels = c(-1.8))
text(x = 1.5, y = -1, labels = c(0.63))
# X2 < 2 with X2 >= 1
lines(x = c(-2, 2), y = c(2, 2))
text(x = 0, y = 2.5, labels = c(2.49))
# X1 < 0 with X2<2 and X2>=1
lines(x = c(0, 0), y = c(1, 2))
text(x = -1, y = 1.5, labels = c(-1.06))
text(x = 1, y = 1.5, labels = c(0.21))

```
## ISL Exercise 8.4.5 (10pts)

Suppose we produce ten bootstrapped samples from a data set
containing red and green classes. We then apply a classifcation tree
to each bootstrapped sample and, for a specifc value of X, produce
10 estimates of P(Class is Red|X):

0.1, 0.15, 0.2, 0.2, 0.55, 0.6, 0.6, 0.65, 0.7, and 0.75.

There are two common ways to combine these results together into a
single class prediction. One is the majority vote approach discussed in
this chapter. The second approach is to classify based on the average
probability. In this example, what is the final classifcation under each
of these two approaches?

```{r}
# Initilize our vector to analyze
p = c(0.1, 0.15, 0.2, 0.2, 0.55, 0.6, 0.6, 0.65, 0.7, 0.75)

# First looking at majority approach

# Checks if the sum of probabilities that are > 50% that class is Red
# is greater than the sum of probabilities that are < 50% that class is Red
sum(p >= 0.5) > sum(p < 0.5)

# Looking at mean approach
mean(p) 
```
The final classification for the majority approach will be red, since
the number of red predictions is greater than the number of green predictions.

The final classifiication for the mean approach is green, since the average of
the probabilites is less than the 50% threshold.

## ISL Lab 8.3. `Boston` data set (30pts)

Follow the machine learning workflow to train regression tree, random forest, and boosting methods for predicting `medv`. Evaluate out-of-sample performance on a test set.

## ISL Lab 8.3 `Carseats` data set (30pts)

Follow the machine learning workflow to train classification tree, random forest, and boosting methods for classifying `Sales <= 8` versus `Sales > 8`. Evaluate out-of-sample performance on a test set.


