---
title: "Biostat 212a Homework 1"
subtitle: "Due Jan 28, 2025 @ 11:59PM"
author: "Nathaniel Boyle and 006525153"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
engine: knitr
knitr:
  opts_chunk: 
    fig.align: 'center'
    # fig.width: 6
    # fig.height: 4
    message: FALSE
    cache: false
---

## Filling gaps in lecture notes (10% pts)

Consider the regression model
$$
Y = f(X) + \epsilon,
$$
where $\operatorname{E}(\epsilon) = 0$. 

### Optimal regression function

Show that the choice
$$
f_{\text{opt}}(X) = \operatorname{E}(Y | X)
$$
minimizes the mean squared prediction error
$$
\operatorname{E}\{[Y - f(X)]^2\},
$$
where the expectations averages over variations in both $X$ and $Y$. (Hint: condition on $X$.)




### Bias-variance trade-off

Given an estimate $\hat f$ of $f$, show that the test error at a $x_0$ can be decomposed as
$$
\operatorname{E}\{[y_0 - \hat f(x_0)]^2\} = \underbrace{\operatorname{Var}(\hat f(x_0)) + [\operatorname{Bias}(\hat f(x_0))]^2}_{\text{MSE of } \hat f(x_0) \text{ for estimating } f(x_0)} + \underbrace{\operatorname{Var}(\epsilon)}_{\text{irreducible}},
$$
where the expectation averages over the variability in $y_0$ and $\hat f$.

## ISL Exercise 2.4.3 (10% pts)

We now revisit the bias-variance decomposition.
(a) Provide a sketch of typical (squared) bias, variance, training error, test error, and Bayes (or irreducible) error curves, on a single plot, as we go from less fexible statistical learning methods
towards more fexible approaches.

The x-axis should represent
the amount of fexibility in the method, and the y-axis should
represent the values for each curve. There should be fve curves.
Make sure to label each one.
(b) Explain why each of the five curves has the shape displayed in
part (a).

```{r, eval = F}
#install.packages("tidyverse")
library(tidyverse)
fit <- lm(sales ~ TV, data = )
```

## ISL Exercise 2.4.4 (10% pts)

You will now think of some real-life applications for statistical learning.

(a) Describe three real-life applications in which classifcation might
be useful. Describe the response, as well as the predictors. Is the
goal of each application inference or prediction? Explain your
answer.

(b) Describe three real-life applications in which regression might
be useful. Describe the response, as well as the predictors. Is the
goal of each application inference or prediction? Explain your
answer.

(c) Describe three real-life applications in which cluster analysis
might be useful.


## ISL Exercise 2.4.10 (30% pts)

This exercise involves the Boston housing data set.
(a) To begin, load in the Boston data set. The Boston data set is
part of the ISLR2 library.

Your can read in the `boston` data set directly from url <https://raw.githubusercontent.com/ucla-biostat-212a/2024winter/master/slides/data/Boston.csv>. A documentation of the `boston` data set is [here](https://www.rdocumentation.org/packages/ISLR2/versions/1.3-2/topics/Boston)

> library(ISLR2)
Now the data set is contained in the object Boston.
> Boston
Read about the data set:
> ?Boston
How many rows are in this data set? How many columns? What
do the rows and columns represent?
(b) Make some pairwise scatterplots of the predictors (columns) in
this data set. Describe your fndings.
(c) Are any of the predictors associated with per capita crime rate?
If so, explain the relationship.
(d) Do any of the census tracts of Boston appear to have particularly
high crime rates? Tax rates? Pupil-teacher ratios? Comment on
the range of each predictor.
(e) How many of the census tracts in this data set bound the Charles
river?
(f) What is the median pupil-teacher ratio among the towns in this
data set?
(g) Which census tract of Boston has lowest median value of owneroccupied homes? What are the values of the other predictors
for that census tract, and how do those values compare to the
overall ranges for those predictors? Comment on your fndings.
(h) In this data set, how many of the census tracts average more than
seven rooms per dwelling? More than eight rooms per dwelling?
Comment on the census tracts that average more than eight
rooms per dwelling.


::: {.panel-tabset}

#### R

```{r, evalue = F}
library(tidyverse)

Boston <- read_csv("https://raw.githubusercontent.com/ucla-biostat-212a/2024winter/master/slides/data/Boston.csv", col_select = -1) %>% 
  print(width = Inf)
```

#### Python

```{python}
import pandas as pd
import io
import requests

url = "https://raw.githubusercontent.com/ucla-econ-425t/2023winter/master/slides/data/Boston.csv"
s = requests.get(url).content
Boston = pd.read_csv(io.StringIO(s.decode('utf-8')), index_col = 0)
Boston
```


:::

## ISL Exercise 3.7.3 (20% pts)

Suppose we have a data set with fve predictors, X1 = GPA, X2 =
IQ, X3 = Level (1 for College and 0 for High School), X4 = Interaction between GPA and IQ, and X5 = Interaction between GPA and
Level. The response is starting salary after graduation (in thousands
of dollars). Suppose we use least squares to ft the model, and get
βˆ0 = 50, βˆ1 = 20, βˆ2 = 0.07, βˆ3 = 35, βˆ4 = 0.01, βˆ5 = −10.

(a) Which answer is correct, and why?

i. For a fxed value of IQ and GPA, high school graduates earn
more, on average, than college graduates.

ii. For a fxed value of IQ and GPA, college graduates earn
more, on average, than high school graduates.

iii. For a fxed value of IQ and GPA, high school graduates earn
more, on average, than college graduates provided that the
GPA is high enough.

iv. For a fxed value of IQ and GPA, college graduates earn
more, on average, than high school graduates provided that
the GPA is high enough.

(b) Predict the salary of a college graduate with IQ of 110 and a
GPA of 4.0.

(c) True or false: Since the coeffcient for the GPA/IQ interaction
term is very small, there is very little evidence of an interaction
efect. Justify your answer.

## ISL Exercise 3.7.15 (20% pts)

This problem involves the Boston data set, which we saw in the lab
for this chapter. We will now try to predict per capita crime rate
using the other variables in this data set. In other words, per capita
crime rate is the response, and the other variables are the predictors.

(a) For each predictor, ft a simple linear regression model to predict
the response. Describe your results. In which of the models is
there a statistically signifcant association between the predictor
and the response? Create some plots to back up your assertions.

(b) Fit a multiple regression model to predict the response using
all of the predictors. Describe your results. For which predictors
can we reject the null hypothesis H0 : βj = 0?

(c) How do your results from (a) compare to your results from (b)?

Create a plot displaying the univariate regression coeffcients
from (a) on the x-axis, and the multiple regression coeffcients
from (b) on the y-axis. That is, each predictor is displayed as a
single point in the plot. Its coeffcient in a simple linear regression model is shown on the x-axis, and its coeffcient estimate
in the multiple linear regression model is shown on the y-axis.

(d) Is there evidence of non-linear association between any of the
predictors and the response? To answer this question, for each
predictor X, ft a model of the form
Y = β0 + β1X + β2X2 + β3X3 + ϵ.

## Bonus question (20% pts)

For multiple linear regression, show that $R^2$ is equal to the correlation between the response vector $\mathbf{y} = (y_1, \ldots, y_n)^T$ and the fitted values $\hat{\mathbf{y}} = (\hat y_1, \ldots, \hat y_n)^T$. That is
$$
R^2 = 1 - \frac{\text{RSS}}{\text{TSS}} = [\operatorname{Cor}(\mathbf{y}, \hat{\mathbf{y}})]^2.
$$

