---
title: "Biostat 212a Homework 1"
subtitle: "Due Jan 28, 2025 @ 11:59PM"
author: "Nathaniel Boyle and 006525153"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
engine: knitr
knitr:
  opts_chunk: 
    fig.align: 'center'
    # fig.width: 6
    # fig.height: 4
    message: FALSE
    cache: false
---

## Filling gaps in lecture notes (10% pts)

Consider the regression model
$$
Y = f(X) + \epsilon,
$$
where $\operatorname{E}(\epsilon) = 0$. 

### Optimal regression function

Show that the choice
$$
f_{\text{opt}}(X) = \operatorname{E}(Y | X)
$$
minimizes the mean squared prediction error
$$
\operatorname{E}\{[Y - f(X)]^2\},
$$
where the expectations averages over variations in both $X$ and $Y$. (Hint: condition on $X$.)




### Bias-variance trade-off

Given an estimate $\hat f$ of $f$, show that the test error at a $x_0$ can be decomposed as
$$
\operatorname{E}\{[y_0 - \hat f(x_0)]^2\} = \underbrace{\operatorname{Var}(\hat f(x_0)) + [\operatorname{Bias}(\hat f(x_0))]^2}_{\text{MSE of } \hat f(x_0) \text{ for estimating } f(x_0)} + \underbrace{\operatorname{Var}(\epsilon)}_{\text{irreducible}},
$$
where the expectation averages over the variability in $y_0$ and $\hat f$.

## ISL Exercise 2.4.3 (10% pts)

We now revisit the bias-variance decomposition.
(a) Provide a sketch of typical (squared) bias, variance, training error, test error, and Bayes (or irreducible) error curves, on a single plot, as we go from less flexible statistical learning methods
towards more flexible approaches.

The x-axis should represent
the amount of flexibility in the method, and the y-axis should
represent the values for each curve. There should be five curves.
Make sure to label each one.
(b) Explain why each of the five curves has the shape displayed in
part (a).

```{r, eval = F}
#install.packages("tidyverse")
library(tidyverse)
fit <- lm(sales ~ TV, data = )
```

## ISL Exercise 2.4.4 (10% pts)

You will now think of some real-life applications for statistical learning.

(a) Describe three real-life applications in which classification might
be useful.Describe the response, as well as the predictors. Is the
goal of each application inference or prediction? Explain your
answer.

One useful example would be email spam detection. Having the ability to quickly sort through important emails and emails that will waste your time is a very practical for daily time management.

The response in this case would be if the email itself is classified as spam or not. This could be a binary classification 1 = spam, 0 = not spam.

The predictors in this case would be features in the emails themselves such as email address, certain keywords that advertisers use, metadata of the email itself (for example what time the email was sent).

The goal in this case is prediction. The model will be introduced to new emails constantly and will need to make a new prediction if the incoming email is classified as spam or not, based on prior data patterns or other parameters set by the model.

Another real-life application would be medical diagnosis. Certain cells in the human body could be classified as cancerous or benign. This is another example of binary classification.

The response in this case could be "cancerous" or benign"

Predictors would be various bio-statistical data pertaining to each patient such as age, blood levels, certain diets, concentration of exotic substances (radiation, carcinogenic chemicals)

The goal in this case would again be prediction. We are taking trained data from previous patients that had confirmed classified cases of cancer or not, and then applying that learned model to new patient data to predict if this new patient has cancer or not.

Last example given here could be in evaluating the effectiveness of a new drug.

The response would be whether or not a patient has improved or not improved.

Predictors in this case could be the dosage amount, overall health of the patients and other various demographics of the individuals that participate in the placebo (control) group vs that of the new drug (experimental) group.

This would be an example of an inference. We are taking in new data and making an inference on how statistically significant is the new drug performing. We are not making any predictive outcomes for individuals that take this drug.


(b) Describe three real-life applications in which regression might
be useful. Describe the response, as well as the predictors. Is the
goal of each application inference or prediction? Explain your
answer.

One example that was brought up in our lectures was advertising and the effect it has on sales.

The response here would the be sales figures themselves (how many units were sold, gross revenue).

The predictors in this case would be the amount spent on advertising, the advertising contents (what is being advertised and how), what time of the year it is (some products will sell better in the Winter for example), and the price of the product itself.

The goal here could depend on what we are trying to achieve. If we want to know how well a future product will do, we could be making a regression prediction. If we want to understand how well advertising expenditure effects previous products we would be doing a inference on that relationship (for example advertising revue vs sales revenue).

Another real-life example would be in housing prices.

The response would be the price of the house.

Predictors would be various features of that house such as number of bedrooms, square footage, location, or other special extensions of the house like a pool.

The goal here is prediction. We are aiming to predict the selling price of a house based on certain criteria. 

Final example here would be for analyzing employee salaries.

The response would be the employee salary.

Predictors would be years of experience, education level, job position, location, an performance statistics if available.

The goal here is to infer how various factors influence an employee's salary. We are aiming to understand the relationship between various factors instead of trying to make a specific prediction of an individual's salary.

(c) Describe three real-life applications in which cluster analysis
might be useful.

One real-life example would be Image segmentation in computer vision projects. Often times we want to divide an image into different segments to help identify what is in a picture for example. There are a wide range of useful reasons to do this, one quick example is diagnosing diseases based of x-ray images or MRI scans by quickly sorting what is ordinary in the scan and what is abnormal.

We could apply cluster analysis to customers in marketing as well. Grouping customers into clusters based of purchasing behavior or demographics can help companies identify various types of customers such as frequent shoppers or one-time buyers. These companies can than look into specifics per group and adjust various tactics such as targeted marking.

Final example could be used for streaming services.
Viewers of certain streaming services like Netflix can be clustered based on factors like minutes watched per day, total viewing sessions per week, or number of unique shows watched per month. These users can be clustered into high or low usage users and then the streaming companies can focus specific marketing strategies on the clusters that are more likely to return the most revenue.

## ISL Exercise 2.4.10 (30% pts)

This exercise involves the Boston housing data set.
(a) To begin, load in the Boston data set. The Boston data set is
part of the ISLR2 library.

```{r, evalue = F}
library(tidyverse)

Boston <- read_csv("https://raw.githubusercontent.com/ucla-biostat-212a/2024winter/master/slides/data/Boston.csv", col_select = -1) %>% 
  print(width = Inf)
```

Read about the data set:
```{r, evalue = F}
#Opens MASS documentation on this Boston dataset
??Boston
```

How many rows are in this data set? How many columns? What
do the rows and columns represent?
506 Rows and 13 columns.
Columns represent various features of a house such as crime rate, average number of rooms per dwelling, mean distances to different employment centers, etc.
The rows represent each individual house.


(b) Make some pairwise scatterplots of the predictors (columns) in
this data set. Describe your findings.

```{r, evalue = F}
#Load MASS library
library(MASS)
#Create simple pairwise scatterplots, -13 to exclude medv which is our response
pairs(Boston[, -13], main = "Pairwise Scatterplots of Predictors")
```
```{r, evalue = F}
#pairwise between distances to employment and crime
plot(Boston$crim, Boston$rm, main = "Scatter Plot of 'crim' vs 'rm'",
     xlab = "Crime Rate (crim)", ylab = "Average Number of Rooms (rm)")
#pairwise between distances to employment and crime
plot(Boston$crim, Boston$dis, main = "Scatter Plot of 'crim' vs 'dis'",
     xlab = "Crime Rate (crim)", ylab = "Average Number of Rooms (rm)")
```

(c) Are any of the predictors associated with per capita crime rate?
If so, explain the relationship.

```{r, evalue = F}
# Calculate correlation between per capita crime rate by town and all other predictors
correlation_matrix <- cor(Boston)
cor_crim <- correlation_matrix["crim",-13]
cor_crim
plot(Boston$crim, Boston$indus, main = "'Crime rate' vs 'non-retail business acres per town.'",
     xlab = "Crime Rate (crim)", ylab = "Average Number of Rooms (rm)")
```
From a simple correlation matrix we can see a few positive correlations with crime per capita. "indus" the proportion of non-retail business acres per town, "nox" which is nitrogen oxides concentration (parts per 10 million), "age" "rad" "tax" "ptratio" and "lstat"

(d) Do any of the census tracts of Boston appear to have particularly
high crime rates? Tax rates? Pupil-teacher ratios?

Comment on the range of each predictor.

```{r, evalue = F}
# Compare the ranges of crime rate, tax rate, and pupil-teacher ratio
range(Boston$crim)
range(Boston$tax)
range(Boston$ptratio)
# Pairwise scatter plot
pairs(Boston[, c("crim", "tax", "ptratio")], main="Pairwise Scatter Plots")
```

(e) How many of the census tracts in this data set bound the Charles
river?
```{r, evalue = F}
# Count how many census tracts are bound by the Charles River
sum(Boston$chas == 1)
```
35 Census bound the Charles river.

(f) What is the median pupil-teacher ratio among the towns in this
data set?

```{r, evalue = F}
# Calculate the median pupil-teacher ratio
median(Boston$ptratio)
```
The median ratio is 19.05.

(g) Which census tract of Boston has lowest median value of owner occupied homes?

```{r, evalue = F}

# Find the index of the census tract with the lowest median value of owner-occupied homes
lowest_medv_index <- which.min(Boston$medv)
# Display the details of the census tract with the lowest median value
Boston[lowest_medv_index, ]
```
Census tract 399 has the lowest median value (5).

What are the values of the other predictors
for that census tract?

From the code printed we see the following values for each predictor:

crim: 38.3518
zn: 0
indus: 18.1
chas: 0
nox: 0.693
rm: 5.453
age: 100
dis: 1.4896	
rad: 24	
tax: 666
ptratio: 20.2
lstat: 30.59

How do those values compare to the
overall ranges for those predictors?

```{r, evalue = F}
# Apply the range function to each predictor in the Boston dataset (excluding the target variable 'medv')
range_df <- sapply(Boston, range)
# Print the range of all predictors
range_df
```

Comment on your findings.

Crime in this census is on the low end
Relatively high levels of nitrogen oxides concentration.
Number of rooms is about on par with the range of rooms in this dataset.
Oldest age compared to dataset.
One of the closest to five employment centers.
Higher end of tax per $10,000
Very high ptratio value.
Very high lstat value.

(h) In this data set, how many of the census tracts average more than
seven rooms per dwelling? More than eight rooms per dwelling?
Comment on the census tracts that average more than eight
rooms per dwelling.

```{r, evalue = F}
# Count how many census tracts have more than 7 rooms per dwelling
more_than_seven_rooms <- sum(Boston$rm > 7)

# Print the result
more_than_seven_rooms

# Count how many census tracts have more than 8 rooms per dwelling
more_than_eight_rooms <- sum(Boston$rm > 8)

# Print the result
more_than_eight_rooms

# Subset the census tracts with more than 8 rooms per dwelling
more_than_eight_df <- Boston[Boston$rm > 8, ]

# View the subset of census tracts with more than 8 rooms
(more_than_eight_df)
```
64 census tracts average more than seven rooms.
13 average more than eight rooms.

## ISL Exercise 3.7.3 (20% pts)

Suppose we have a data set with five predictors, X1 = GPA, X2 =
IQ, X3 = Level (1 for College and 0 for High School), X4 = Interaction between GPA and IQ, and X5 = Interaction between GPA and
Level. The response is starting salary after graduation (in thousands
of dollars). Suppose we use least squares to ft the model, and get
βˆ0 = 50, βˆ1 = 20, βˆ2 = 0.07, βˆ3 = 35, βˆ4 = 0.01, βˆ5 = −10.

(a) Which answer is correct, and why?

iv. For a fixed value of IQ and GPA, college graduates earn
more, on average, than high school graduates provided that
the GPA is high enough.

Because there is a high positive coefficient attracted to the level of education,
our model suggests that the level of education (being in college) is a strong
predictor in having a higher salary (from βˆ3 = 35). The negative coefficient
(βˆ5 = −10) for the interaction between GPA and level of education, suggests
that a certain level of GPA is needed for the correlation of salary and 
education level to be visualized. 


(b) Predict the salary of a college graduate with IQ of 110 and a
GPA of 4.0.

$$ Salary=50+20⋅GPA+0.07⋅IQ+35⋅Level+0.01⋅(GPA×IQ)+(−10)⋅(GPA×Level)$$

Plug in values
$$ Level = 1, GPA = 4.0, IQ = 110$$

$$ Salary=50+80+7.7+35+4.4−40=137.1 $$

(c) True or false: Since the coefficient for the GPA/IQ interaction
term is very small, there is very little evidence of an interaction
effect. Justify your answer.

False. We would need to perform a null-hypothesis test or report the standard error
on this coefficient in order to evaluate how significant this value is.


## ISL Exercise 3.7.15 (20% pts)

This problem involves the Boston data set, which we saw in the lab
for this chapter. We will now try to predict per capita crime rate
using the other variables in this data set. In other words, per capita
crime rate is the response, and the other variables are the predictors.

(a) For each predictor, ft a simple linear regression model to predict
the response. Describe your results. In which of the models is
there a statistically significant association between the predictor
and the response? Create some plots to back up your assertions.

```{r, evalue = F}
# Load the necessary libraries
library(MASS)

# Get the list of predictors (exclude 'CRIM' from the dataset)
predictors <- setdiff(names(Boston), "crim")

# Loop through each predictor and fit the linear model
models <- lapply(predictors, function(predictor) {
  lm(crim ~ get(predictor), data = Boston)
})

# Display summaries for each model
model_summaries <- lapply(models, summary)
print(model_summaries)

# Create the scatter plot with the regression line

plot_list <- lapply(predictors, function(predictor) {
  ggplot(Boston, aes_string(x = predictor, y = "crim")) +
    geom_point() +  # scatter plot
    geom_smooth(method = "lm", se = FALSE, color = "blue") +  # regression line
    labs(title = paste("Crime Rate vs", predictor),
         x = predictor, 
         y = "Per Capita Crime Rate (criim)") +
    theme_minimal()
})

# Print plots to check the result
print(plot_list[])


```

(b) Fit a multiple regression model to predict the response using
all of the predictors. Describe your results. For which predictors
can we reject the null hypothesis H0 : βj = 0?

```{r, evalue = F}

# Fit the multiple regression model using all predictors
multiple_model <- lm(crim ~ ., data = Boston)

# Display the summary of the model
summary(multiple_model)
```
From this list of predictors we can reject the null for the following predictors:
zn
dis
rad
medv

(c) How do your results from (a) compare to your results from (b)?

Create a plot displaying the univariate regression coefficients
from (a) on the x-axis, and the multiple regression coefficients
from (b) on the y-axis. That is, each predictor is displayed as a
single point in the plot. Its coefficient in a simple linear regression model is shown on the x-axis, and its coefficient estimate
in the multiple linear regression model is shown on the y-axis.

(d) Is there evidence of non-linear association between any of the
predictors and the response? To answer this question, for each
predictor X, ft a model of the form
Y = β0 + β1X + β2X2 + β3X3 + ϵ.

## Bonus question (20% pts)

For multiple linear regression, show that $R^2$ is equal to the correlation between the response vector $\mathbf{y} = (y_1, \ldots, y_n)^T$ and the fitted values $\hat{\mathbf{y}} = (\hat y_1, \ldots, \hat y_n)^T$. That is
$$
R^2 = 1 - \frac{\text{RSS}}{\text{TSS}} = [\operatorname{Cor}(\mathbf{y}, \hat{\mathbf{y}})]^2.
$$

