---
title: "Biostat 212a Homework 2"
subtitle: "Due Feb 8, 2025 @ 11:59PM"
author: "Nathaniel Boyle and 006525153"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
engine: knitr
knitr:
  opts_chunk: 
    fig.align: 'center'
    # fig.width: 6
    # fig.height: 4
    message: FALSE
    cache: false
---

## ISL Exercise 4.8.1 (10pts)

Using a little bit of algebra, prove that (4.2) is equivalent to (4.3). In
other words, the logistic function representation and logit representation for the logistic regression model are equivalent.

$$p(X) = \frac{e^{\beta_0 + \beta_1 X}}{1 + e^{\beta_0 + \beta_1 X}}$$

$$\frac{p(X)} {1 − p(X)} = eβ_0+β_1X$$
We have $p(X) = \frac{e^{\beta_0 + \beta_1 X}}{1 + e^{\beta_0 + \beta_1 X}}$ by definition
Thus,
$$ 1-p(X)= 1- \frac{e^{\beta_0 + \beta_1 X}}{1 + e^{\beta_0 + \beta_1 X}}$$
Simplify with common denominator

$$ 1- \frac{e^{\beta_0 + \beta_1 X}}{1 + e^{\beta_0 + \beta_1 X}} = \frac{1 + e^{\beta_0 + \beta_1 X}-e^{\beta_0 + \beta_1 X}}{1 + e^{\beta_0 + \beta_1 X}} = \frac{1}{1 + e^{\beta_0 + \beta_1 X}} = 1-p(X)$$
Therefore,
$$\frac{p(X)}{1-p(X)} = \frac{(1 + e^{\beta_0 + \beta_1 X})e^{\beta_0 + \beta_1 X}}{1 + e^{\beta_0 + \beta_1 X}}=e^{\beta_0 + \beta_1 X}$$
## ISL Exercise 4.8.6 (10pts)
Suppose we collect data for a group of students in a statistics class
with variables $X_1 = hours$ studied, $X_2 = undergrad GPA$, 
and Y =receive an A. We ft a logistic regression and produce estimated
coefficient, $\hatβ_0 = −6$, $\hatβ_1 = 0.05$, $\hatβ_2 = 1$.

Estimate the probability that a student who studies for 40 h and
has an undergrad GPA of 3.5 gets an A in the class

From our logistic regression model we plug in our values to obtain a logit value:
$$logit(p) = \hatβ_0 + \hatβ_1X_1 +\hatβ_2X_2 = −6 + 0.05(40) + 1(3.5) = -0.5$$
Now we convert this logit value to a probability:
$$ p = \frac{1}{1+e^{-logit(p)}} = \frac{1}{1+e^{0.5}} ≈ 0.378 $$
This means our student in this case has a roughly 37.8% chance of getting an A.

How many hours would the student in part (a) need to study to
have a 50 % chance of getting an A in the class?

A 50% chance is equivalent to $logit(p)=0$ thus we can solve for X, the
amount of hours we study for:

$$0 = −6 + 0.05(X) + 1(3.5)$$
$$\frac{6-3.5}{0.05} = X = 50$$
Thus a student with a GPA of 3.5 would need 50 hours of studying for a 50% chance
to get an A, according to this logistic model.

## ISL Exercise 4.8.9 (10pts)

This problem has to do with odds.

(a) On average, what fraction of people with an odds of 0.37 of
defaulting on their credit card payment will in fact default?

The odds of something occurring is expressed as a ratio of the probability
that the event will occur to the probability that it will not occur

$$\frac{p}{1-p}=odds$$
Given our odds are 0.37, we can solve for p with some algebra.
$$ p=0.37(1-p) = 0.37-0.37p $$
$$1.37p = 0.37$$
$$p=\frac{0.37}{1.37}≈0.27 $$
Thus the fraction of people with an odds of 0.37 that default will be roughly 
27% of people.

(b) Suppose that an individual has a 16 % chance of defaulting on
her credit card payment. What are the odds that she will default?

Here we have the probability $p=0.16$, thus we simply solve for the ratio:
$$\frac{p}{1-p} = odds = \frac{0.16}{1-0.16} ≈ 0.19$$
Thus the odds of an individual with a 16% chance of defaulting is 0.19.
## ISL Exercise 4.8.13 (a)-(i) (50pts)

This question should be answered using the Weekly data set, which
is part of the ISLR2 package. This data is similar in nature to the
Smarket data from this chapter’s lab, except that it contains 1, 089
weekly returns for 21 years, from the beginning of 1990 to the end of
2010.

(a) Produce some numerical and graphical summaries of the Weekly
data.
```{r}
# Load the ISLR2 package
#install.packages("ISLR2")
library(ISLR2)

#documentation
?Weekly

# Show the first few rows of the dataset
head(Weekly)

#print statistical summary
summary(Weekly)

#correlation matrix
cor(Weekly[, -9])  # Excluding 'Direction' because it's categorical

#histogram of Volume
hist(Weekly$Volume, main="Histogram of Volume", xlab="Volume", col="lightblue", border="black")

# Boxplot of 'Today' returns by 'Direction' (categorical variable)
boxplot(Today ~ Direction, data=Weekly, main="Boxplot of Today's Returns by Direction", 
        xlab="Direction", ylab="Today's Returns", col=c("lightgreen", "lightcoral"))

# Scatter plot between 'Lag1' and 'Lag2' to check their relationship
plot(Weekly$Lag1, Weekly$Lag2, main="Scatter Plot of Lag1 vs Lag2",
     xlab="Lag1", ylab="Lag2", pch=19, col=rgb(0,0,1,0.5))
#Scatter plot between 'Lag1' and 'Lag3' to check their relationship
plot(Weekly$Lag1, Weekly$Lag3, main="Scatter Plot of Lag1 vs Lag3",
     xlab="Lag1", ylab="Lag2", pch=19, col=rgb(0,0,1,0.5))
```
 Do there appear to be any patterns?
 
 From our correlation matrix, we can see that volume of trades and year is
 strongly correlated. This indicates seasonal changes in the market that
 heavily impact trades.
 
 The lags are not very strongly correlated with much of anything in this dataset.
 There could still be a relationship worth exploring, but nothing from first
 glance suggests a strong correlation or auto-correlation.
 
 Our "direction" column is visualizing returns on "up" days and "down" days.
 We see one outlier on the "down" which may indicate the market crash.
```{r}
#
min_return <- min(Weekly$Today)  # Find the minimum return value

# Find the index of the row with the minimum return
min_return_index <- which(Weekly$Today == min_return)

# Get the year corresponding to the row with the minimum return
min_return_year <- Weekly$Year[min_return_index]

# Print the result
min_return_year
 
```
 
 The "volume" is heavily right-skewed. This means most days there was very
 little trading occurring, but there was a period of time
 when a large amount of trading suddenly occurred.
 This can occur during market crashes. In fact,
 this dataset covers the 2008 financial crisis.
 
```{r}
# Filter data for the year 2008
data_2008 <- subset(Weekly, Year == 2008)

# Summarize the data for 2008
summary(data_2008$Volume)
summary(Weekly$Volume)

# Create a histogram for the trading volume in 2008 and overlay them
# Plot the first histogram for 2008 trading volume
hist(data_2008$Volume, main = "Overlayed Histogram of Trading Volume",
     xlab = "Volume (in billions)", col = rgb(0.6, 0.8, 1, 0.5), # Light blue with transparency
     border = "black", xlim = range(c(data_2008$Volume, Weekly$Volume)), # Same x-axis limits
     ylim = c(0, 100), # Adjust the y-axis if necessary to fit both histograms
     breaks = 20)  # Control the number of bins

# Overlay the second histogram for the entire dataset trading volume
hist(Weekly$Volume, col = rgb(0.2, 0.8, 0.2, 0.5), border = "black", 
     add = TRUE, breaks = 20)  # `add = TRUE` adds the histogram to the existing plot

# Add a legend to distinguish the two histograms
legend("topright", legend = c("2008 Volume", "All Data Volume"),
       fill = c(rgb(0.6, 0.8, 1, 0.5), rgb(0.2, 0.8, 0.2, 0.5)), bty = "n")


# You could also look at the boxplot to identify any outlines
# Set up the plotting area to display two plots side by side
par(mfrow = c(1, 2)) 

# Boxplot for trading volume in 2008
boxplot(data_2008$Volume, main = "2008 Trading Volume", ylab = "Volume (in billions)")

# Boxplot for the entire dataset's trading volume
boxplot(Weekly$Volume, main = "All Trading Volume", ylab = "Volume (in billions)")

# Reset the plotting layout to default (one plot per graph)
par(mfrow = c(1, 1)) 

```
We can see from the Volume in 2008 and the average Volume of the dataset,
the average ("mean") volume of trades was nearly 5 times higher in 2008 than the
rest of the dataset.

Looking at the boxplots for 2008 Trading volume as well
confirms that a rather unusual amount of trading occurred. This can probably be
attributed to the 2008 financial crisis we discussed before.

The histogram of the volume is also confirms this higher than usual trading
visually.

(b) Use the full data set to perform a logistic regression with
Direction as the response and the five lag variables plus Volume
as predictors.
Use the summary function to print the results. 
```{r}
# 2. Fit the logistic regression model with the five lag variables and Volume as predictors
model <- glm(Direction ~ Lag1 + Lag2 + Lag3 + Lag4 + Lag5 + Volume, 
             data = Weekly, 
             family = binomial)

# 3. View the summary of the logistic regression model
summary(model)
```
Do any of the predictors appear to be statistically significant? If so,
which ones?

From the model lag 2 is the only significant predictor.

(c) Compute the confusion matrix and overall fraction of correct
predictions. Explain what the confusion matrix is telling you
about the types of mistakes made by logistic regression.

(d) Now ft the logistic regression model using a training data period
from 1990 to 2008, with Lag2 as the only predictor. Compute the
confusion matrix and the overall fraction of correct predictions
for the held out data (that is, the data from 2009 and 2010).

(e) Repeat (d) using LDA.

(f) Repeat (d) using QDA.

(g) Repeat (d) using KNN with K = 1.

(h) Repeat (d) using naive Bayes.

(i) Which of these methods appears to provide the best results on
this data?

## Bonus question: ISL Exercise 4.8.13 Part (j) (30pts)

## Bonus question: ISL Exercise 4.8.4 (30pts)
