---
title: "Biostat 212A Homework 3"
subtitle: "Due Feb 18, 2025 @ 11:59PM"
author: "Nathaniel Boyle and 006525153"
date: today
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
engine: knitr
knitr:
  opts_chunk: 
    fig.align: 'center'
    # fig.width: 6
    # fig.height: 4
    message: FALSE
    cache: false
---

## ISL Exercise 5.4.2 (10pts)

We will now derive the probability that a given observation is part
of a bootstrap sample. Suppose that we obtain a bootstrap sample
from a set of n observations.

(a) What is the probability that the first bootstrap observation is
not the jth observation from the original sample? Justify your
answer.

Total Observations: There are 
$n$ observations in the original sample.

When drawing the first observation for the bootstrap sample, each of the $n$
observations has an equal probability of being chosen.

The probability that the first bootstrap observation is the j-th observation is:$$𝑃=\frac{1}{n}$$

Thus the probability of not selecting that j-th observation is $$P=1-\frac{1}{n}$$ 

(b) What is the probability that the second bootstrap observation is not the jth observation from the original sample?

Since the bootstrap sampling is done with replacement, the probability distribution of the second observation is identical to the first. 
 
Thus the probability that the second bootstrap observation is not from the original
sample is: $$P=1-\frac{1}{n}$$

This is the same as the probability for the first observation.

(c) Argue that the probability that the jth observation is not in the
bootstrap sample is $(1 − 1/n)^n$.

Since each bootstrap observation is done with replacement, and since we have
already shown in part (a) and (b) that the probability is $P=1-\frac{1}{n}$
we can extrapolate this result to all draws of $n$

Thus
$$P=(1-\frac{1}{n})^{n}$$

(d) When n = 5, what is the probability that the jth observation is
in the bootstrap sample?

Plug in our value for $n=5$ and recognize we want the probability of the bootstrap
being the same as the observation, thus we will do $1-P(j*)$ where $P(j*)$ is the probability
of **not** getting the j-th observation in the bootstrap sample.
Thus,

$$P(j)=1-P(j*)=1-(1-\frac{1}{n})^{n}$$
Plug in values for $n$

$$P(j)=1-(1-\frac{1}{5})^5=0.67232$$
Therefore our probability that the jth observation is in the bootstrap sample
when $n=5$ is about $67.23\%$.

(e) When n = 100, what is the probability that the jth observation
is in the bootstrap sample?

Same steps, now with $n=100$

$$P(j)=1-(1-\frac{1}{100})^{100}=0.6339$$


(f) When n = 10, 000, what is the probability that the jth observation is in the
bootstrap sample?

Same steps, now with $n=100$

$$P=1-(1-\frac{1}{10000})^{10000}=0.63213$$

(g) Create a plot that displays, for each integer value of n from 1
to 100, 000, the probability that the jth observation is in the
bootstrap sample. Comment on what you observe.

```{r}
# create a plot that shows the probability of getting bootstrap sample from n=1
# to n=100 

# Define the function to calculate the probability
probability_in_bootstrap <- function(n) {
  return(1 - (1 - 1/n)^n)
}

# Generate a sequence of n from 1 to 100,000
n_values <- 1:100000

# Calculate the probabilities for each n
probabilities <- sapply(n_values, probability_in_bootstrap)

# Plot the result
plot(n_values, probabilities, type = "l", col = "blue", lwd = 2,
     xlab = "n (Sample size)", ylab = "Probability the j-th observation is in the bootstrap sample",
     main = "Probability that the j-th observation is in the bootstrap sample")
grid()

```

We observe that the probability curve takes an 'L' shape. After a few sample sizes
the probability sharply stabilizes around $62\%$ and increasing the sample size
drastically has no observable effect to reduce the probability.

(h) We will now investigate numerically the probability that a bootstrap sample
of size n = 100 contains the jth observation. Here
j = 4. We repeatedly create bootstrap samples, and each time
we record whether or not the fourth observation is contained in
the bootstrap sample.
```{r}
store <- rep(NA, 10000)
for(i in 1:10000){
store[i] <- sum(sample(1:100, rep=TRUE) == 4) > 0
 }

mean(store)
```
We observe that on average we get the 4th observation of our bootstrap sample
is about $62\%$ which is roughly what we observed in part (g).

## ISL Exercise 5.4.9 (20pts)

We will now consider the Boston housing data set, from the ISLR2
library.

(a) Based on this data set, provide an estimate for the population
mean of medv. Call this estimate µˆ.

```{r}
# Load the ISLR2 library
library(ISLR2)

# Calculate the sample mean of medv
mean_medv <- mean(Boston$medv)

# Print the result
mean_medv
```

(b) Provide an estimate of the standard error of µˆ. Interpret this
result.
```{r}
# Calculate the standard deviation of medv
sd_medv <- sd(Boston$medv)

# Get the number of observations in the dataset
n <- nrow(Boston)

# Calculate the standard error of the mean
se_medv <- sd_medv / sqrt(n)

# Print the result
se_medv
```
Hint: We can compute the standard error of the sample mean by
dividing the sample standard deviation by the square root of the
number of observations.

(c) Now estimate the standard error of µˆ using the bootstrap. How
does this compare to your answer from (b)?
```{r}
# Set a seed for reproducibility
set.seed(123)

# Define the number of bootstrap samples
n_bootstrap <- 1000

# Initialize a vector to store bootstrap means
bootstrap_means <- numeric(n_bootstrap)

# Perform the bootstrap procedure
for (i in 1:n_bootstrap) {
  # Resample the data with replacement
  bootstrap_sample <- sample(Boston$medv, size = n, replace = TRUE)
  
  # Compute the mean of the bootstrap sample
  bootstrap_means[i] <- mean(bootstrap_sample)
}

# Estimate the standard error using the bootstrap
se_bootstrap <- sd(bootstrap_means)

# Print the bootstrap standard error
se_bootstrap
```

(d) Based on your bootstrap estimate from (c), provide a 95 % confdence interval for the mean of medv. Compare it to the results
obtained using t.test(Boston$medv).
Hint: You can approximate a 95 % confdence interval using the
formula [ˆµ − 2SE(ˆµ), µˆ + 2SE(ˆµ)].
```{r}
# Calculate the sample mean of medv (from part a)
mean_medv <- mean(Boston$medv)

# Construct the 95% confidence interval using the bootstrap estimate of SE
lower_bound <- mean_medv - 2 * se_bootstrap
upper_bound <- mean_medv + 2 * se_bootstrap

# Print the 95% confidence interval
lower_bound
upper_bound

# Perform a t-test to calculate the confidence interval
t_test_result <- t.test(Boston$medv)

# Print the 95% confidence interval from the t-test
t_test_result$conf.int
```

(e) Based on this data set, provide an estimate, µˆmed, for the median
value of medv in the population.

```{r}
# Calculate the sample median of medv
median_medv <- median(Boston$medv)

# Print the result
median_medv

```

(f) We now would like to estimate the standard error of µˆmed. Unfortunately, there is no simple formula for computing the standard
error of the median. Instead, estimate the standard error of the
median using the bootstrap. Comment on your fndings.

```{r}
# Set a seed for reproducibility
set.seed(123)

# Define the number of bootstrap samples
n_bootstrap <- 1000

# Initialize a vector to store bootstrap medians
bootstrap_medians <- numeric(n_bootstrap)

# Perform the bootstrap procedure
for (i in 1:n_bootstrap) {
  # Resample the data with replacement
  bootstrap_sample <- sample(Boston$medv, size = n, replace = TRUE)
  
  # Compute the median of the bootstrap sample
  bootstrap_medians[i] <- median(bootstrap_sample)
}

# Estimate the standard error using the bootstrap
se_median_bootstrap <- sd(bootstrap_medians)

# Print the bootstrap standard error for the median
se_median_bootstrap

```

(g) Based on this data set, provide an estimate for the tenth percentile of medv in Boston census tracts. Call this quantity µˆ0.1.
(You can use the quantile() function.)

```{r}
# Calculate the 10th percentile of medv
percentile_10th <- quantile(Boston$medv, 0.1)

# Print the result
percentile_10th

```

(h) Use the bootstrap to estimate the standard error of µˆ0.1. Comment on your fndings.


```{r}
# Set a seed for reproducibility
set.seed(123)

# Define the number of bootstrap samples
n_bootstrap <- 1000

# Initialize a vector to store bootstrap 10th percentiles
bootstrap_percentiles_10th <- numeric(n_bootstrap)

# Perform the bootstrap procedure
for (i in 1:n_bootstrap) {
  # Resample the data with replacement
  bootstrap_sample <- sample(Boston$medv, size = n, replace = TRUE)
  
  # Compute the 10th percentile of the bootstrap sample
  bootstrap_percentiles_10th[i] <- quantile(bootstrap_sample, 0.1)
}

# Estimate the standard error using the bootstrap
se_percentile_10th_bootstrap <- sd(bootstrap_percentiles_10th)

# Print the bootstrap standard error for the 10th percentile
se_percentile_10th_bootstrap

```

## Least squares is MLE (10pts)

Show that in the case of linear model with Gaussian errors, maximum likelihood and least squares are the same thing, and $C_p$ and AIC are equivalent.


## ISL Exercise 6.6.1 (10pts)

## ISL Exercise 6.6.3 (10pts)

## ISL Exercise 6.6.4 (10pts)

## ISL Exercise 6.6.5 (10pts)

## ISL Exercise 6.6.11 (30pts)

You must follow the [typical machine learning paradigm](https://ucla-biostat-212a.github.io/2024winter/slides/06-modelselection/workflow_lasso.html) to compare _at least_ 3 methods: least squares, lasso, and ridge. Report final results as

| Method | CV RMSE | Test RMSE |
|:------:|:------:|:------:|:------:|
| LS | | | |
| Ridge | | | |
| Lasso | | | |
| ... | | | |

## Bonus question (20pts)

Consider a linear regression, fit by least squares to a set of training data $(x_1, y_1), \ldots, (x_N,  y_N)$ drawn at random from a population. Let $\hat \beta$ be the least squares estimate. Suppose we have some test data $(\tilde{x}_1, \tilde{y}_1), \ldots, (\tilde{x}_M, \tilde{y}_M)$ drawn at random from the same population as the training data. If $R_{\text{train}}(\beta) = \frac{1}{N} \sum_{i=1}^N (y_i - \beta^T x_i)^2$ and $R_{\text{test}}(\beta) = \frac{1}{M} \sum_{i=1}^M (\tilde{y}_i - \beta^T \tilde{x}_i)^2$. Show that
$$
\operatorname{E}[R_{\text{train}}(\hat{\beta})] < \operatorname{E}[R_{\text{test}}(\hat{\beta})].
$$