---
title: "Biostat 212a Homework 6"
subtitle: "Due Mar 22, 2025 @ 11:59PM"
author: "Nathaniel Boyle and 006525153"
date: today
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
engine: knitr
knitr:
  opts_chunk: 
    fig.align: 'center'
    # fig.width: 6
    # fig.height: 4
    message: FALSE
    cache: false
    warning: false
---

Load R libraries.
```{r}
if (!requireNamespace("pheatmap", quietly = TRUE)) {
  install.packages("pheatmap", repos = "https://cloud.r-project.org/")
}
if (!requireNamespace("ggdendro", quietly = TRUE)) {
  install.packages("ggdendro", repos = "https://cloud.r-project.org/")
}
if (!requireNamespace("tidyclust", quietly = TRUE)) {
  install.packages("tidyclust", repos = "https://cloud.r-project.org/")
}
if (!requireNamespace("RcppHungarian", quietly = TRUE)) {
  install.packages("RcppHungarian", repos = "https://cloud.r-project.org/")
}
if (!requireNamespace("glmnet", quietly = TRUE)) {
  install.packages("glmnet", repos = "https://cloud.r-project.org/")
}
if (!requireNamespace("dplyr", quietly = TRUE)) {
  install.packages("dplyr", repos = "https://cloud.r-project.org/")
}

library(tidyverse)
library(tidymodels)
library(readr)
library(tswge)
library(ggplot2)
library(ggdendro)
library(glmnet)
library(yardstick)
library(workflows)
library(parsnip)
library(tidyclust)
library(RcppHungarian)
library(xgboost)
library(caret)
library(dplyr)
library(randomForest)
library(stats)
library(reshape2)
library(gridExtra)
library(pheatmap)

```

## New York Stock Exchange (NYSE) data (1962-1986) (140 pts)

Historical trading statistics from the New York Stock Exchange. Daily values of the normalized log trading volume, DJIA return, and log volatility are shown for a 24-year period from 1962-1986. We wish to predict trading volume on any day, given the history on all earlier days. To the left of the red bar (January 2, 1980) is training data, and to the right test data.


The [`NYSE.csv`](https://raw.githubusercontent.com/ucla-biostat-212a/2025winter/master/slides/data/NYSE.csv) file contains three daily time series from the New York Stock Exchange (NYSE) for the period Dec 3, 1962-Dec 31, 1986 (6,051 trading days).

- `Log trading volume` ($v_t$): This is the fraction of all outstanding shares that are traded on that day, relative to a 100-day moving average of past turnover, on the log scale.
    
- `Dow Jones return` ($r_t$): This is the difference between the log of the Dow Jones Industrial Index on consecutive trading days.
    
- `Log volatility` ($z_t$): This is based on the absolute values of daily price movements.

```{r}
# Read in NYSE data from url

url = "https://raw.githubusercontent.com/ucla-biostat-212a/2025winter/master/slides/data/NYSE.csv"
NYSE <- read_csv(url)

NYSE
```
```{r}
acfdf <- function(vec) {
  vacf <- acf(vec, plot = F)
  with(vacf, data.frame(lag, acf))
}

ggacf <- function(vec) {
  ac <- acfdf(vec)
  ggplot(data = ac, aes(x=lag, y = acf)) +geom_hline(aes(yintercept = 0)) +
    geom_segment(mapping = aes(xend = lag, yend = 0))
}

tplot <- function(vec) {
  df <- data.frame(X = vec, t = seq_along(vec))
  ggplot(data = df, aes(x = t, y = X)) +geom_line()
}


```
The **autocorrelation** at lag $\ell$ is the correlation of all pairs $(v_t, v_{t-\ell})$ that are $\ell$ trading days apart. These sizable correlations give us confidence that past values will be helpful in predicting the future.

```{r}
#| code-fold: true
#| label: fig-nyse-autocor
#| fig-cap: "The autocorrelation function for log volume. We see that nearby values are fairly strongly correlated, with correlations above 0.2 as far as 20 days apart."

ggacf(NYSE$log_volume) + ggthemes::theme_few()

```

Do a similar plot for (1) the correlation between $v_t$ and lag $\ell$ `Dow Jones return` $r_{t-\ell}$ and (2) correlation between $v_t$ and lag $\ell$ `Log volatility` $z_{t-\ell}$.

```{r}
seq(1, 30) %>% 
  map(function(x) {cor(NYSE$log_volume , lag(NYSE$DJ_return, x),
                       use = "pairwise.complete.obs")}) %>% 
  unlist() %>% 
  tibble(lag = 1:30, cor = .) %>% 
  ggplot(aes(x = lag, y = cor)) + 
  geom_hline(aes(yintercept = 0)) + 
  geom_segment(mapping = aes(xend = lag, yend = 0)) + 
  ggtitle("AutoCorrelation between `log volume` and lagged `DJ return`")
```

```{r}
seq(1, 30) %>% 
  map(function(x) {cor(NYSE$log_volume , lag(NYSE$log_volatility, x),
                       use = "pairwise.complete.obs")}) %>% 
  unlist() %>% 
  tibble(lag = 1:30, cor = .) %>% 
  ggplot(aes(x = lag, y = cor)) + 
  geom_hline(aes(yintercept = 0)) + 
  geom_segment(mapping = aes(xend = lag, yend = 0)) + 
  ggtitle("AutoCorrelation between `log volume` and lagged `log volatility`")
```



### Project goal

Our goal is to forecast daily `Log trading volume`, using various machine learning algorithms we learnt in this class. 

The data set is already split into train (before Jan 1st, 1980, $n_{\text{train}} = 4,281$) and test (after Jan 1st, 1980, $n_{\text{test}} = 1,770$) sets.

<!-- Include `day_of_week` as a predictor in the models. -->

In general, we will tune the lag $L$ to acheive best forecasting performance. In this project, we would fix $L=5$. That is we always use the previous five trading days' data to forecast today's `log trading volume`.

Pay attention to the nuance of splitting time series data for cross validation. Study and use the [`time-series`](https://www.tidymodels.org/learn/models/time-series/) functionality in tidymodels. Make sure to use the same splits when tuning different machine learning algorithms.

Use the $R^2$ between forecast and actual values as the cross validation and test evaluation criterion.

### Baseline method (20 pts)

We use the straw man (use yesterday’s value of `log trading volume` to predict
that of today) as the baseline method.
Evaluate the $R^2$ of this method on the test data.

```{r}
#baseline method

# Create the baseline prediction (using the previous day's log trading volume)
NYSE_baseline <- NYSE |>
  mutate(BaselinePrediction = lag(log_volume, n = 1)) |>
  filter(!is.na(BaselinePrediction))
  
# View the first few rows to check the baseline predictions
head(NYSE_baseline)

# Check the structure of the data to ensure that the Date column is correctly parsed
str(NYSE_baseline)
head(NYSE_baseline)

# Convert 'Date' column to Date type (if not already)
NYSE_baseline$Date <- as.Date(NYSE_baseline$date, format = "%m/%d/%Y")

# Ensure that the 'log_volume' column exists and is named correctly
head(NYSE_baseline)

# Create the baseline prediction (using the previous day's log trading volume)
NYSE_baseline <- NYSE_baseline |>
  mutate(BaselinePrediction = lag(log_volume, n = 1)) |>
  filter(!is.na(BaselinePrediction))

# View the first few rows to check the baseline predictions
head(NYSE_baseline)

# Split the data into training and test sets (before and after January 1st, 1980)
train_data <- NYSE_baseline %>%
  filter(Date < "1980-01-01")

test_data <- NYSE_baseline %>%
  filter(Date >= "1980-01-01")

# View the train and test set sizes to ensure proper split
dim(train_data)
dim(test_data)

# Calculate the R^2 for the baseline model on the test data
# Use lm (linear model) to calculate R^2 between predicted and actual log trading volume
model <- lm(log_volume ~ BaselinePrediction, data = test_data)

# Get the R^2 value
summary(model)$r.squared
```

### Autoregression (AR) forecaster (30 pts)

- Let
$$
y = \begin{pmatrix} v_{L+1} \\ v_{L+2} \\ v_{L+3} \\ \vdots \\ v_T \end{pmatrix}, \quad M = \begin{pmatrix}
1 & v_L & v_{L-1} & \cdots & v_1 \\
1 & v_{L+1} & v_{L} & \cdots & v_2 \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
1 & v_{T-1} & v_{T-2} & \cdots & v_{T-L}
\end{pmatrix}.
$$

- Fit an ordinary least squares (OLS) regression of $y$ on $M$, giving
$$
\hat v_t = \hat \beta_0 + \hat \beta_1 v_{t-1} + \hat \beta_2 v_{t-2} + \cdots + \hat \beta_L v_{t-L},
$$
known as an **order-$L$ autoregression** model or **AR($L$)**.

```{r}
# Example time series data
log_volume <- NYSE$log_volume

# Define the order of the autoregression (L=3 for AR(3) model)
L <- 3

# Create the matrix M with lagged values (first column is for the intercept term)
T <- length(log_volume)
M <- cbind(
  1,  # intercept term (constant)
  lag(log_volume, 1)[(L+1):T],  # lag 1
  lag(log_volume, 2)[(L+1):T],  # lag 2
  lag(log_volume, 3)[(L+1):T]   # lag 3
)

# Create the vector y (the target variable, which is log_volume from L+1 to T)
y <- log_volume[(L+1):T]

# Fit the OLS model using the lm() function (exclude intercept in the formula since it's already included in M)
model <- lm(y ~ M - 1)  # '-1' removes the intercept term (since it's already in M)

# View the model coefficients
summary(model)

```

- Tune AR(5) with elastic net (lasso + ridge) regularization using all 3 features on the training data, and evaluate the test performance. 

```{r}

# Convert Date column to Date type if it's not already
NYSE$date <- as.Date(NYSE$date, format = "%m/%d/%Y")

# Create lagged features for log_volume, DJIA_return, and log_volatility
L <- 5  # We will use AR(5) model, so L = 5

NYSE_lagged <- NYSE %>%
  mutate(
    lag_log_volume_1 = lag(log_volume, 1),
    lag_log_volume_2 = lag(log_volume, 2),
    lag_log_volume_3 = lag(log_volume, 3),
    lag_log_volume_4 = lag(log_volume, 4),
    lag_log_volume_5 = lag(log_volume, 5),
    
    lag_DJ_return_1 = lag(DJ_return, 1),
    lag_DJ_return_2 = lag(DJ_return, 2),
    lag_DJ_return_3 = lag(DJ_return, 3),
    lag_DJ_return_4 = lag(DJ_return, 4),
    lag_DJ_return_5 = lag(DJ_return, 5),
    
    lag_log_volatility_1 = lag(log_volatility, 1),
    lag_log_volatility_2 = lag(log_volatility, 2),
    lag_log_volatility_3 = lag(log_volatility, 3),
    lag_log_volatility_4 = lag(log_volatility, 4),
    lag_log_volatility_5 = lag(log_volatility, 5)
  ) %>%
  filter(complete.cases(.))  # Remove rows with NA values caused by lagging

# Split data into training (before 1980) and test (after 1980)
train_data <- NYSE_lagged %>% filter(date < "1980-01-01")
test_data <- NYSE_lagged %>% filter(date >= "1980-01-01")

# Inspect the sizes of the train and test sets
dim(train_data)
dim(test_data)


# Prepare training data
X_train <- as.matrix(train_data %>% select(starts_with("lag_")))  # All lagged features
y_train <- train_data$log_volume  # Target variable

# Prepare test data
X_test <- as.matrix(test_data %>% select(starts_with("lag_")))
y_test <- test_data$log_volume


# Fit the Elastic Net model using cross-validation
cv_model <- cv.glmnet(X_train, y_train, alpha = 0.5, nfolds = 10)  # alpha = 0.5 for Elastic Net

# View the cross-validation results
cv_model

# Plot the cross-validation curve
plot(cv_model)


# Predict on the test set
y_pred <- predict(cv_model, X_test, s = "lambda.min")

# Calculate R^2 on the test set
rss <- sum((y_test - y_pred)^2)
tss <- sum((y_test - mean(y_test))^2)
r_squared <- 1 - (rss / tss)

# Print the R^2 value
r_squared

```

- Hint: [Workflow: Lasso](https://ucla-biostat-212a.github.io/2025winter/slides/06-modelselection/workflow_lasso.html) is a good starting point.

### Random forest forecaster (30pts)

- Use the same features as in AR($L$) for the random forest. Tune the random forest and evaluate the test performance.

```{r}

# Create lagged features for log_volume, DJIA_return, and log_volatility
L <- 5  # We will use AR(5) model, so L = 5

NYSE_lagged <- NYSE %>%
  mutate(
    lag_log_volume_1 = lag(log_volume, 1),
    lag_log_volume_2 = lag(log_volume, 2),
    lag_log_volume_3 = lag(log_volume, 3),
    lag_log_volume_4 = lag(log_volume, 4),
    lag_log_volume_5 = lag(log_volume, 5),
    
    lag_DJ_return_1 = lag(DJ_return, 1),
    lag_DJ_return_2 = lag(DJ_return, 2),
    lag_DJ_return_3 = lag(DJ_return, 3),
    lag_DJ_return_4 = lag(DJ_return, 4),
    lag_DJ_return_5 = lag(DJ_return, 5),
    
    lag_log_volatility_1 = lag(log_volatility, 1),
    lag_log_volatility_2 = lag(log_volatility, 2),
    lag_log_volatility_3 = lag(log_volatility, 3),
    lag_log_volatility_4 = lag(log_volatility, 4),
    lag_log_volatility_5 = lag(log_volatility, 5)
  ) %>%
  filter(complete.cases(.))  # Remove rows with NA values caused by lagging

# Split data into training (before 1980) and test (after 1980)
train_data <- NYSE_lagged %>% filter(date < "1980-01-01")
test_data <- NYSE_lagged %>% filter(date >= "1980-01-01")

# Inspect the sizes of the train and test sets
dim(train_data)
dim(test_data)

# Prepare the training data matrix and target variable
X_train <- train_data %>% select(starts_with("lag_"))
y_train <- train_data$log_volume

# Prepare the test data matrix and target variable
X_test <- test_data %>% select(starts_with("lag_"))
y_test <- test_data$log_volume

# Set up a training control object for tuning (use 10-fold cross-validation)
train_control <- trainControl(method = "cv", number = 10)

# Correct the tuning grid: Only include mtry for Random Forest
tune_grid <- expand.grid(
  mtry = c(2, 4, 6, 8, 10)  # Number of variables to consider at each split
)

# Tune the random forest model
rf_tune <- train(
  x = X_train, 
  y = y_train,
  method = "rf", # Random Forest method
  trControl = train_control,
  tuneGrid = tune_grid, 
  ntree = 500  # Number of trees in the forest
)

# View the best tuned parameters
print(rf_tune$bestTune)

#our best forest model was with mtry 6
# this means 6 variables considered each split provided
# the best outcome

```
```{r}
#evaluate model perofmrance

# Make predictions on the test set
y_pred_rf <- predict(rf_tune, X_test)

# Calculate R^2 on the test set
rss_rf <- sum((y_test - y_pred_rf)^2)
tss_rf <- sum((y_test - mean(y_test))^2)
r_squared_rf <- 1 - (rss_rf / tss_rf)

# Print the R^2 value
print(r_squared_rf)

#found R^2 is 0.4, lower than previous models.

```

- Hint: [Workflow: Random Forest for Prediction](https://ucla-biostat-212a.github.io/2025winter/slides/08-tree/workflow_rf_reg.html) is a good starting point.

### Boosting forecaster (30pts)

- Use the same features as in AR($L$) for the boosting. Tune the boosting algorithm and evaluate the test performance.

```{r, warning=FALSE}

# Create lagged features for log_volume, DJIA_return, and log_volatility
L <- 5  # We will use AR(5) model, so L = 5

NYSE_lagged <- NYSE %>%
  mutate(
    lag_log_volume_1 = lag(log_volume, 1),
    lag_log_volume_2 = lag(log_volume, 2),
    lag_log_volume_3 = lag(log_volume, 3),
    lag_log_volume_4 = lag(log_volume, 4),
    lag_log_volume_5 = lag(log_volume, 5),
    
    lag_DJ_return_1 = lag(DJ_return, 1),
    lag_DJ_return_2 = lag(DJ_return, 2),
    lag_DJ_return_3 = lag(DJ_return, 3),
    lag_DJ_return_4 = lag(DJ_return, 4),
    lag_DJ_return_5 = lag(DJ_return, 5),
    
    lag_log_volatility_1 = lag(log_volatility, 1),
    lag_log_volatility_2 = lag(log_volatility, 2),
    lag_log_volatility_3 = lag(log_volatility, 3),
    lag_log_volatility_4 = lag(log_volatility, 4),
    lag_log_volatility_5 = lag(log_volatility, 5)
  ) %>%
  filter(complete.cases(.))  # Remove rows with NA values caused by lagging

# Split data into training (before 1980) and test (after 1980)
train_data <- NYSE_lagged %>% filter(date < "1980-01-01")
test_data <- NYSE_lagged %>% filter(date >= "1980-01-01")

# Prepare the training data matrix and target variable
X_train <- train_data %>% select(starts_with("lag_"))
y_train <- train_data$log_volume

# Prepare the test data matrix and target variable
X_test <- test_data %>% select(starts_with("lag_"))
y_test <- test_data$log_volume

# Convert the data to matrix format for xgboost
X_train_matrix <- as.matrix(X_train)
X_test_matrix <- as.matrix(X_test)

# Set up a training control object for tuning (use 5-fold cross-validation)
train_control <- trainControl(method = "cv", number = 5)

# Set up the correct tuning grid
tune_grid <- expand.grid(
  nrounds = c(10, 50, 100),           # boosting iterations
  max_depth = c(3, 5),             # Limit max_depth
  eta = c(0.01, 0.1),              # Use smaller eta values
  gamma = c(0, 0.1),               # Reduce gamma values
  colsample_bytree = c(0.7),       # Keep colsample_bytree constant
  subsample = c(0.7),              # Keep subsample constant
  min_child_weight = c(1)          # Reduce min_child_weight values
)


# Train the xgboost model with tuning
xgb_tune <- train(
  x = X_train_matrix,
  y = y_train,
  method = "xgbTree",  # Use xgboost algorithm
  trControl = train_control,
  tuneGrid = tune_grid
)

# View the best-tuned parameters
print(xgb_tune$bestTune)


# Make predictions on the test set
y_pred_xgb <- predict(xgb_tune, X_test_matrix)

# Calculate R^2 on the test set
rss_xgb <- sum((y_test - y_pred_xgb)^2)
tss_xgb <- sum((y_test - mean(y_test))^2)
r_squared_xgb <- 1 - (rss_xgb / tss_xgb)

# Print the R^2 value
print(paste("R^2 for xgboost model:", r_squared_xgb))


```

- Hint: [Workflow: Boosting tree for Prediction](https://ucla-biostat-212a.github.io/2025winter/slides/08-tree/workflow_boosting_reg.html) is a good starting point.


### Summary (30pts)

Your score for this question is largely determined by your final test performance.

Summarize the performance of different machine learning forecasters in the following format. 

```{r}
# Extract CV R² from the Random Forest model
print("Random Forest Cross-Validation R² values:")
print(rf_tune$results[, c("mtry", "RMSE", "Rsquared")])

# Predict on the test set
rf_predictions <- predict(rf_tune, newdata = test_data)

# Calculate test R²
rf_test_r2 <- cor(rf_predictions, test_data$log_volume)^2
print(paste("Random Forest Test R²:", rf_test_r2))

# Extract CV R² from the XGBoost model
print("XGBoost Cross-Validation R² values:")
print(xgb_tune$results[, c("nrounds", "max_depth", "eta", "Rsquared")])

# Predict on the test set
xgb_predictions <- predict(xgb_tune, newdata = test_data)

# Calculate test R²
xgb_test_r2 <- cor(xgb_predictions, test_data$log_volume)^2
print(paste("XGBoost Test R²:", xgb_test_r2))

```

| Method | CV $R^2$ | Test $R^2$ |
|:------:|:------:|:------:|:------:|
| Baseline | | | |
| AR(5) | | | |
| Random Forest | | | |
| Boosting | | | |

## ISL Exercise 12.6.13 (90 pts)
On the book website, www.statlearning.com, there is a gene expression data set (Ch12Ex13.csv) that consists of 40 tissue samples with
measurements on 1,000 genes. The frst 20 samples are from healthy
patients, while the second 20 are from a diseased group.

(a) Load in the data using read.csv(). You will need to select
header = F.

```{r}
#load dataset

# URL to the CSV file
url <- "http://www.statlearning.com/s/Ch12Ex13.csv"

# Load the dataset directly from the URL, ignore header
data <- read.csv(url, header = F)

# Check the first few rows of the data
head(data)

```


(b) Apply hierarchical clustering to the samples using correlationbased
distance, and plot the dendrogram.

```{r}

# Compute the correlation matrix
correlation_matrix <- cor(t(data))

# Convert the correlation matrix into a distance matrix (1 - correlation)
dist_matrix <- as.dist(1 - correlation_matrix)

# Perform hierarchical clustering
hc <- hclust(dist_matrix)

# Convert the hierarchical clustering object to a data frame suitable for ggplot
dendrogram_data <- as.dendrogram(hc)

# Plot the dendrogram using ggplot2
ggdendrogram(dendrogram_data, rotate = FALSE, theme_dendro = TRUE) + 
  labs(title = "Hierarchical Clustering Dendrogram", 
       x = "Samples", 
       y = "Height") + 
  theme_minimal() + 
  theme(panel.background = element_blank(),  # Remove the grey background
        axis.text.x = element_text(size = 12, angle =45),    # Adjust axis text size if needed
        axis.title = element_text(size = 14)) +
  scale_x_continuous(breaks = seq(0, nrow(data), by = 50))  # Display every 100th tick# Adjust axis title size

```

Do the genes separate
the samples into the two groups?

From what it looks like we can see at least two distinct groups around sample
No. 600 to 750.

```{r}
#different linkage methods

# Create a list of methods for hierarchical clustering
methods <- c("complete", "single", "average", "ward.D", "centroid")

# Function to create a dendrogram plot for each method
create_dendrogram <- function(method, dist_matrix) {
  # Perform hierarchical clustering with the specified method
  hc <- hclust(dist_matrix, method = method)
  
  # Convert the hierarchical clustering object to a data frame suitable for ggplot
  dendrogram_data <- as.dendrogram(hc)
  
  # Plot the dendrogram using ggplot2
  p <- ggdendrogram(dendrogram_data, rotate = FALSE, theme_dendro = TRUE) + 
    labs(title = paste(method), 
         x = "Samples", 
         y = "Height") + 
    theme_minimal() + 
    theme(panel.background = element_blank(),  # Remove the grey background
          axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate only x-axis labels
          axis.title = element_text(size = 14)) +  # Adjust axis title size
    scale_x_continuous(breaks = seq(0, nrow(data), by = 100))  # Display every 100th tick
  
  return(p)
}

# Create a list of dendrogram plots for each method
dendrogram_plots <- lapply(methods, create_dendrogram, dist_matrix = dist_matrix)

# Arrange the plots in a grid
grid.arrange(grobs = dendrogram_plots, ncol = 2)  # 2 columns for the layout


```


Do your results depend on the
type of linkage used?

It still appears that there are 2 distinct groups of some sort present in all the
different linkage types. However how these groups are presented are each a
little different.

### PCA and UMAP (30 pts)

Now we want to see the clusters from the genes using a PCA map
and a UMAP.

```{r}
# PCA Maps

# Standardize the data (center and scale)
data_scaled <- scale(data)

# Perform PCA
pca_result <- prcomp(data_scaled)

# Extract the PCA results (first two components)
pca_data <- data.frame(pca_result$x)

# Assuming you have some known groupings (e.g., based on hierarchical clustering), 
# you can assign cluster labels to each sample (e.g., for two groups).
# You can use hierarchical clustering to decide the groups or use a known column.
# Example: Let's assume you want to color points by hierarchical clustering:

# Perform clustering again with a chosen method (e.g., complete linkage)
hc <- hclust(dist_matrix, method = "complete")

# Cut the tree into two clusters
clusters <- cutree(hc, k = 2)  # Assuming you want 2 clusters

# Add the cluster labels to the PCA data
pca_data$Cluster <- factor(clusters)

# Plot the PCA results using ggplot2
library(ggplot2)
ggplot(pca_data, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(size = 3) +
  labs(title = "PCA of Gene Expression Data",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal() +
  scale_color_manual(values = c("blue", "red"))  # Customize the colors as needed

```
```{r}
# UMAP
# Load the necessary library for UMAP
library(umap)

# Perform UMAP on the scaled data
umap_result <- umap(data_scaled)

# Extract UMAP results (first two components)
umap_data <- data.frame(UMAP1 = umap_result$layout[, 1], 
                        UMAP2 = umap_result$layout[, 2])

# Assuming you have cluster labels from hierarchical clustering (as before)
umap_data$Cluster <- factor(clusters)  # Use the clusters from hierarchical clustering

# Plot the UMAP results using ggplot2
library(ggplot2)
ggplot(umap_data, aes(x = UMAP1, y = UMAP2, color = Cluster)) +
  geom_point(size = 3) +
  labs(title = "UMAP of Gene Expression Data",
       x = "UMAP 1",
       y = "UMAP 2") +
  theme_minimal() +
  scale_color_manual(values = c("blue", "red"))  # Customize the colors as needed

```
```{r}
# Test block

transposed_data <- data |>
  t()|>
  as.data.frame()|>
  as_tibble()|>
  rename(ID = rowname)


names(transposed_data) <- c("ID", paste(seq(1,1000),""))

transposed_data <- transposed_data |>
  mutate(status = as.character(rep(c(1,0), each = 20)), .after = "ID")

```

### 12.6.13 (c) (30 pts)

Your collaborator wants to know which genes difer the most
across the two groups. Suggest a way to answer this question,
and apply it here.

What we can first do is cut our dendrogram to form
two groups. These groups will correspond to the
experminatl groups we are intersted in comparing.

We can perform a differential expression analysis (t-test) for each gene.

For each gene a t-test is performed comparing
the expression values between the two groups.
The t-test checks if there is a significant difference in gene expression 
between the two groups for each individual gene. The null hypothesis 
is that the mean expression level is the same in both groups, 
and the alternative hypothesis is that they are different.
We then iterate over each row (gene) of the transposed data and run 
the t-test for each gene.


Since you're testing many genes at once, there's a risk of obtaining false 
positives due to multiple comparisons. 
The Benjamini-Hochberg (BH) procedure adjusts the p-values to control
the false discovery rate (FDR).

The p.adjust() function applies this correction to the original p-values
to account for the fact that you are testing many hypotheses at once.
```{r}

# Transpose the data to cluster on samples
data_transposed <- t(data)

# Compute the correlation distance (1 - correlation)
cor_dist <- 1 - cor(data_transposed)

# Perform hierarchical clustering using the "complete" linkage method
hclust_res <- hclust(as.dist(cor_dist), method = "complete")

# Cut the dendrogram to form two groups
groups <- cutree(hclust_res, k = 2)  # Two groups

# Perform differential expression analysis (t-test) for each gene
p_values <- apply(data_transposed, 1, function(gene_expression) {
  # Perform t-test between the two groups for each gene
  t_test_result <- t.test(gene_expression ~ groups)
  return(t_test_result$p.value)
})

# Adjust p-values for multiple testing (e.g., using the Benjamini-Hochberg procedure)
adjusted_p_values <- p.adjust(p_values, method = "BH")

# Create a data frame of results
results <- data.frame(
  Gene = rownames(data_transposed),
  P_Value = p_values,
  Adjusted_P_Value = adjusted_p_values
)

# Sort the results by adjusted p-value to find the most significant genes
results <- results[order(results$Adjusted_P_Value), ]

# Display the top genes that are most different between the two groups
head(results)


```

```{r}

# Select the top 10 most differentially expressed genes based on adjusted p-values
top_genes <- head(results$Gene, 10)

# Extract expression data for these genes
top_gene_data <- data_transposed[top_genes, ]

# Create a heatmap of the top genes
pheatmap(top_gene_data, cluster_rows = TRUE, cluster_cols = TRUE,
         show_rownames = TRUE, show_colnames = TRUE,
         scale = "row", main = "Top 10 Differentially Expressed Genes")
```


