---
title: "Biostat 212a Homework 5"
subtitle: "Due Mar 16, 2024 @ 11:59PM"
author: "Nathaniel Boyle and 006525153"
date: today
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
engine: knitr
knitr:
  opts_chunk: 
    fig.align: 'center'
    # fig.width: 6
    # fig.height: 4
    message: FALSE
    cache: false
---

## ISL Exercise 9.7.1 (10pts)

### 1. Hyperplanes in two dimensions.
#### part a. and part b. 

Sketch the hyperplane 1+3X1−X2=0.
Indicate the set of points for which 1+3X1−X2>0
as well as the set of points for which 1+3X1−X2<0.

On the same plot, sketch the hyperplane −2+X1+2X2=0
Indicate the set of points for which −2+X1+2X2>0
as well as the set of points for which −2+X1+2X2<0

```{r}
library(ggplot2)

# Define the data for the two equations
x1 <- -10:10
x2 <- 1 + 3 * x1
x3 <- 1 - x1 / 2

# Create a data frame for ggplot
df <- data.frame(x1 = x1, x2 = x2, x3 = x3)

# Plot the graph using ggplot2
ggplot(df, aes(x = x1)) +
  # Plot the first line (blue)
  geom_line(aes(y = x2), color = "blue") +
  
  # Annotate text for the blue line
  annotate("text", x = 3, y = -5, label = "Greater than 0", color = "blue") +
  annotate("text", x = -1, y = 15, label = "Less than 0", color = "blue") +
  
  # Plot the second line (red)
  geom_line(aes(y = x3), color = "red") +
  
  # Annotate text for the red line
  annotate("text", x = 3, y = -10, label = "Less than 0", color = "red") +
  annotate("text", x = -1, y = 20, label = "Greater than 0", color = "red") +
  
  # Add the legend with LaTeX formatting
  scale_color_manual(
    values = c("blue", "red"),
    labels = c(
      expression(1 + 3 * x[1] - x[2] == 0),
      expression(-2 + x[1] + 2 * x[2] == 0)
    )
  ) +
  
  # Add labels and title
  labs(
    title = "Equations Plot",
    x = expression(x[1]),
    y = expression(x[2])
  ) +
  
  # Set theme for better appearance
  theme_minimal() +
  theme(legend.position = "right")
```


## ISL Exercise 9.7.2 (10pts)

We have seen that in p = 2 dimensions, a linear decision boundary
takes the form $β_0+β_1X_1+β_2X_2 = 0$. We now investigate a non-linear
decision boundary.

### 2.a and 2.b

Sketch the curve $(1 + X1)^2 + (2 − X2)^2 = 4$.
On your sketch, indicate the set of points for which
$(1 + X1)^2 + (2 − X2)^2 > 4$,
as well as the set of points for which
$(1 + X1)^2 + (2 − X2)^2 ≤ 4$

```{r}
# Load ggplot2 library
library(ggplot2)

# Define the equation of the circle (1 + X1)^2 + (2 - X2)^2 = 4
# The center is (1, 2) and the radius is 2

radius <- 2
center_x <- 1
center_y <- 2

# Generate points to plot the circle
theta <- seq(0, 2*pi, length.out = 100)  # Angle sequence
x1 <- center_x + radius * cos(theta)  # X1 coordinates
x2 <- center_y + radius * sin(theta)  # X2 coordinates

# Create a data frame for ggplot
circle_data <- data.frame(X1 = x1, X2 = x2)

# Create a grid of points to evaluate the inequality
grid_x1 <- seq(-4, 4, by = 0.1)
grid_x2 <- seq(-1, 5, by = 0.1)
grid <- expand.grid(X1 = grid_x1, X2 = grid_x2)

# Evaluate the inequality (1 + X1)^2 + (2 - X2)^2 at each point
grid$distance_squared <- (1 + grid$X1)^2 + (2 - grid$X2)^2

# Plot the circle using ggplot
ggplot() +
  # Shading for (1 + X1)^2 + (2 - X2)^2 < 4 (inside the circle)
  geom_tile(data = subset(grid, distance_squared < 4),
            aes(x = X1, y = X2, fill = "Inside the circle ( < 4)"), 
            color = "white", width = 0.1, height = 0.1) +
  
  # Shading for (1 + X1)^2 + (2 - X2)^2 > 4 (outside the circle)
  geom_tile(data = subset(grid, distance_squared > 4),
            aes(x = X1, y = X2, fill = "Outside the circle ( > 4)"), 
            color = "white", width = 0.1, height = 0.1) +
  
  
  # Add labels and title
  labs(x = "X1", y = "X2",
       title = "Non-Linear Decision Boundary with Inequalities") +
  
  # Theme and legend customization
  theme_minimal() +
  scale_fill_manual(
    values = c("Inside the circle ( < 4)" = "lightblue",
               "Outside the circle ( > 4)" = "lightpink")) +
  
  coord_fixed(ratio = 1) +  # Ensure equal aspect ratio
  theme(legend.position = "right")  # Place legend to the right
```

### 2.c 
Suppose that a classier assigns an observation to the blue class if
$(1 + X1)2 + (2 − X2)2 > 4$,
and to the red class otherwise. To what class is the observation
$(0, 0)$ classified? $(−1, 1)$? $(2, 2)$? $(3, 8)$?

```{r}
# Load ggplot2 library
library(ggplot2)

# Define the equation of the circle (1 + X1)^2 + (2 - X2)^2 = 4
# The center is (1, 2) and the radius is 2

radius <- 2
center_x <- 1
center_y <- 2

# Generate points to plot the circle
theta <- seq(0, 2*pi, length.out = 100)  # Angle sequence
x1 <- center_x + radius * cos(theta)  # X1 coordinates
x2 <- center_y + radius * sin(theta)  # X2 coordinates

# Create a data frame for ggplot
circle_data <- data.frame(X1 = x1, X2 = x2)

# Create a grid of points to evaluate the inequality
grid_x1 <- seq(-4, 4, by = 0.1)
grid_x2 <- seq(-1, 10, by = 0.1)
grid <- expand.grid(X1 = grid_x1, X2 = grid_x2)

# Evaluate the inequality (1 + X1)^2 + (2 - X2)^2 at each point
grid$distance_squared <- (1 + grid$X1)^2 + (2 - grid$X2)^2

# Define the observations to classify
observations <- data.frame(
  X1 = c(0, -1, 2, 3),
  X2 = c(0, 1, 2, 8),
  Label = c("Outside the circle", "Inside the circle",
            "Outside the circle", "Outside the circle")
)

# Plot the circle using ggplot2
ggplot() +
  # Shading for (1 + X1)^2 + (2 - X2)^2 < 4 (inside the circle)
  geom_tile(data = subset(grid, distance_squared < 4),
            aes(x = X1, y = X2, fill = "Inside the circle ( < 4)"), 
            color = "white", width = 0.1, height = 0.1) +
  
  # Shading for (1 + X1)^2 + (2 - X2)^2 > 4 (outside the circle)
  geom_tile(data = subset(grid, distance_squared > 4),
            aes(x = X1, y = X2, fill = "Outside the circle ( > 4)"), 
            color = "white", width = 0.1, height = 0.1) +
  
  # Plot the observations with classifications
  geom_point(data = observations,
             aes(x = X1, y = X2, color = Label), size = 3) +
  
  # Add coordinates as text labels next to the points
  geom_text(data = observations,
            aes(x = X1, y = X2, label = paste("(", X1, ",", X2, ")", sep = "")), 
            vjust = 1.5, hjust = -0.1, size = 4, color = "black") +
  
  # Add labels and title
  labs(x = "X1", y = "X2",
       title = "Non-Linear Decision Boundary with Classifications") +
  
  # Theme and legend customization
  theme_minimal() +
  
  scale_fill_manual(
    values = c("Inside the circle ( < 4)" = "lightblue",
               "Outside the circle ( > 4)" = "lightpink")) +
  scale_color_manual(
    values = c("Outside the circle" = "blue", "Inside the circle" = "red")) +
  
  coord_fixed(ratio = 1) +  # Ensure equal aspect ratio
  theme(legend.position = "right")  # Place legend to the right
```

### 2.d 
Argue that while the decision boundary in (c) is not linear in
terms of $X_1$ and $X_2$, it is linear in terms of
$X_1$, $X_1^2$ , $X_2$, and$X_2^2$.

The decision boundary in the given problem is represented by the equation:
$$(1+X_1)^2 + (2 - X_2)^2 = 4$$
Although this represents a circle, if we transform the variables we can express
this boundary in a linear form.

Expanding our squares to represent our boundary in terms of $X_1$,
$X_1^2$ , $X_2$, and $X_2^2$:

$$(1+2X_1 +X_1^2 )+(4 − 4X_2 + X_2^2)=4$$

Simplify: 

$$2X_1 − 4X_2 + X_1^2 + X_2^2 = −1$$

We can now transform our boundary terms into new, Linear terms of $Z_n$, where
n is an integer that represents a new variable transformation for any $X_n$.

$$Z_1 = X_1$$
$$Z_2 = X_1^2 $$
$$Z_3 = X_2 $$
$$Z_4 = X_2^2 $$


Now our decision boundary becomes:

$$ 2Z_1 − 4Z_3 + Z_2 + Z_4 = −1 $$
This *is* a linear equation. Therefore our decision boundary is linear in terms
of $X_1$, $X_1^2$ , $X_2$, and$X_2^2$ when we transform them to $Z_n$
This is the core of what SVMs do with non-linear boundaries. Transforming the
data into higher-dimensional feature space allows models to apply linear
classifiers to non-linear equations.


## Support vector machines (SVMs) on the `Carseats` data set (30pts)

Follow the machine learning workflow to train support vector classifier (same as SVM with linear kernel), SVM with polynomial kernel (tune the degree and regularization parameter $C$), and SVM with radial kernel (tune the scale parameter $\gamma$ and regularization parameter $C$) for classifying `Sales<=8` versus `Sales>8`. Use the same seed as in your HW4 for the initial test/train split and compare the final test AUC and accuracy to those methods you tried in HW4.

```{r eval=FALSE}
# Install necessary packages if not already installed
install.packages(c("ISLR", "e1071", "caret"))

# Load the libraries
library(ISLR)
library(e1071)  # SVM implementation
library(caret)   # For model training and tuning


# Load the carseats dataset
data("Carseats")

# Create a binary target variable: Sales <= 8 versus Sales > 8
Carseats$SalesBinary <- ifelse(Carseats$Sales <= 8, 0, 1)

# Create a training and testing split
set.seed(123)
trainIndex <- createDataPartition(Carseats$SalesBinary, p = 0.7, list = FALSE)
trainData <- Carseats[trainIndex, ]
testData <- Carseats[-trainIndex, ]



# Standardize the training and testing sets
preProcValues <- preProcess(trainData[, -c(1, 2, 11)], method = c("center", "scale"))
trainDataScaled <- predict(preProcValues, trainData[, -c(1, 2, 11)])
testDataScaled <- predict(preProcValues, testData[, -c(1, 2, 11)])

# Add the target variable back
trainDataScaled$SalesBinary <- trainData$SalesBinary
testDataScaled$SalesBinary <- testData$SalesBinary



# SVM with linear kernel
svm_linear <- svm(SalesBinary ~ ., data = trainDataScaled, kernel = "linear")

# Make predictions
pred_linear <- predict(svm_linear, testDataScaled)

# Evaluate the model
confusionMatrix(pred_linear, testData$SalesBinary)

# Tune SVM with polynomial kernel
tune_poly <- tune(svm, SalesBinary ~ ., data = trainDataScaled, 
                  kernel = "polynomial", 
                  ranges = list(degree = 2:4, cost = 10^(-1:1)))

# Get the best model
svm_poly <- tune_poly$best.model

# Make predictions
pred_poly <- predict(svm_poly, testDataScaled)

# Evaluate the model
confusionMatrix(pred_poly, testData$SalesBinary)



# Tune SVM with radial kernel
tune_radial <- tune(svm, SalesBinary ~ ., data = trainDataScaled, 
                    kernel = "radial", 
                    ranges = list(sigma = seq(0.01, 1, by = 0.1), cost = 10^(-1:1)))

# Get the best model
svm_radial <- tune_radial$best.model

# Make predictions
pred_radial <- predict(svm_radial, testDataScaled)

# Evaluate the model
confusionMatrix(pred_radial, testData$SalesBinary)

# Compare the accuracy of each model
cat("Linear Kernel Accuracy: ", sum(pred_linear == testData$SalesBinary) / length(pred_linear), "\n")
cat("Polynomial Kernel Accuracy: ", sum(pred_poly == testData$SalesBinary) / length(pred_poly), "\n")
cat("Radial Kernel Accuracy: ", sum(pred_radial == testData$SalesBinary) / length(pred_radial), "\n")

```

## Bonus (10pts)

Let
$$
f(X) = \beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p = \beta_0 + \beta^T X. 
$$
Then $f(X)=0$ defines a hyperplane in $\mathbb{R}^p$. Show that $f(x)$ is proportional to the signed distance of a point $x$ to the hyperplane $f(X) = 0$. 
